Symmetric Circuits for Rank Gates

### 2018-02-28 sections/background.tex
Updates to the entire paper 28th Feb 2018

Let $\FP[\tau]$ denote \emph{fixed-point logic} with respect to $\tau$. The
logic $\FP$ is an extension of $\FO$ by an inflationary fixed-point operator. In
other words, $\FP[\tau]$ includes all formulas in $\FO[\tau]$, along with all
formulas of the form $[\ifp_{V, \vec{x}}. \phi]$, where $V$ is a relation symbol
not in $\tau$, $\phi \in \FO[\tau \cup \{V\}]$, and $\vec{x}$ is a sequence of
variables of length $\arty(V)$. We assume the standard syntax and semantics for
$\FO[\tau]$ and $\FP[\tau]$. For more detail on the syntax and semantics of
these logics please see~\cite{grohe2017descriptive} and~\cite{GradelP15a}.

Let $L$ be a logic and $\tau$ a vocabulary. For $\phi \in L[\tau]$ and a
sequence of variables $\vec{x}$, we write $\phi(\vec{x})$ if the free variables
in $\phi$ are among those in $\vec{x}$. Suppose $\vec{x} = (x_1, \ldots , x_k)$,
and let $\mathcal{A}$ be a $\tau$-structure with universe $U$. We call a
function $\alpha : {x_1, \ldots, x_k} \rightarrow U$ an \emph{assignment}. If
$\vec{a}\in U^k$, we write $\alpha^{\vec{x}}_{\vec{a}}$ to denote the assignment
that maps the variable $x_i$ to $a_i$ for $i \in [k]$. We say that two
assignments $\alpha$ and $\beta$ are \emph{compatible} if $\alpha$ and $\beta$
agree on the intersection of their domain. We write $\mathcal{A} \models_L
\phi[\alpha]$ if, and only if, $\phi$ is true in $\mathcal{A}$ with respect to
$L$ where for all $i \in [k]$ each variable $x_i$ is assigned to $\alpha(x_i)$.
If the logic is obvious from context we omit the subscript and just write
$\mathcal{A} \models \phi[\vec{a}]$.

\subsection{Fixed-Point with Counting}
Let $\FPC[\tau]$ denote \emph{fixed-point logic with counting} with respect to
the vocabulary $\tau$. The logic $\FPC$ is an extension of $\FP$ that includes a
counting operator that allows the logic to determine the size of a definable
set, as well as other mechanisms for reasoning about quantities. We now briefly
describe the structure of a formula of $\FPC[\tau]$. Every first-order variable
is assigned a type, and so each variable is either a \emph{vertex} or
\emph{number} variable. Vertex variables are assigned to elements of the
universe of the structure, and play a similar role to the variables that appear
in formulas in $\FO[\tau]$ or $\FP[\tau]$. Number variables are assigned to
elements of $\nats$. We usually use Latin letters (e.g. $x, y, z, \ldots$) to
denote vertex variables and Greek letters (e.g. $\mu, \nu, \epsilon, \ldots$) to
denote number variables. All atomic formulas in $\FP[\tau]$ are atomic formulas
in $\FPC[\tau]$. We say that $t$ is a \emph{number term} if $t$ a number
variable or if $t$ is an application of the \emph{counting operator}, i.e. $t =
\#x. \phi(x)$, where $\phi \in \FPC[\tau]$ and $x$ is a vertex variable free in
$\phi$. If $t_1$ and $t_2$ are number terms then $t_1 \leq t_2$ and $t_1 = t_2$
are atomic formulas. $\FPC[\tau]$ includes all formulas formed by closing the
set of atomic formulas under the usual Boolean connectives, the first-order
quantifiers (for vertex variables), and the inflationary fixed-point operator.
The second-order variables, such as those that appear in an application of the
fixed-point operator, are allowed to have mixed-type, i.e. we associate with a
second-order variable $R$ a type $(k, l) \in \nats^2$ intended to mean that $R$
is assigned to a subset of $U^k \times \nats^l$.

@@

If $\mathcal{A}$ is a $\tau$-structure with universe $U$, the counting term
$\#x. \phi(x)$ evaluates to $\vert \{a \in U : \mathcal{A} \models \phi[a]\}
\vert$ in $\mathcal{A}$, assuming a fixed assignment to the other free variables
in $\phi$. The intended semantics for bounded quantification is that the
variable may only be assigned to a value at most equal to the the evaluation of
the number term bounding the variable. The semantics of the `$=$' and `$\leq$'
symbols are defined for number terms in the obvious manner. For a more detailed
description of the syntax and semantics of $\FPC$ please
see~\cite{grohe2017descriptive} or the introduction of~\cite{GradelP15a}.

@@

Let $\FOC[\tau]$ denote \emph{first-order logic with counting} with respect to
the vocabulary $\tau$, and be the set of all formulas in $\FPC[\tau]$ that do
not contain the an application of the fixed-point operator.

@@

\subsection{Rank Logic}
Let $\FPR[\tau]$ denote \emph{fixed-point logic with rank} (or just \emph{rank
  logic}) with respect to the vocabulary $\tau$. The logic $\FPR$ is an
extension of $\FP$ to include an operator that decides the rank of a definable
matrix over a finite field, as well as other mechanisms for reasoning about
quantity. In this way we can think of the structure of a formula of $\FPR$ as
being similar to that of $\FPC$, with rank operators replacing counting
operators.

We now briefly describe the structure of a formula in $\FPR[\tau]$. For more
detail on the syntax and semantics of $\FPR$ please see~\cite{GradelP15a}
and~\cite{Dawar09logicswith}. As in the case of $\FPC$, each variable is either
a number or vertex variable, with vertex variables being assigned to elements of
the universe and number variables assigned to natural numbers. All atomic
formulas in $\FP[\tau]$ are atomic formulas in $\FPR[\tau]$. We say that $t$ is
a \emph{number term} if $t$ is a number variable or if $t$ is an application of
the \emph{rank operator}, i.e. $t = [\rank (\vec{x}, \vec {\nu} \leq \vec{t},
\vec{y}\vec{\mu} \leq \vec{s}, \pi \leq \eta). \phi]$, where $\phi$ is a number
term, $\vec{t}$ and $\vec{s}$ are tuples of number terms bounding the sequences
of number variables $\vec{\mu}$ and $\vec{\nu}$, and $\eta$ is a number term
bounding the number variable $\pi$. If $t_1$ and $t_2$ are number terms then
$t_1 \leq t_2$ and $t_1 = t_2$ are atomic formulas. The formulas of $\FPR$ are
then formed by closing the set of atomic formulas under the usual Boolean
connectives, the first-order quantifiers, and the fixed-point operator. As in
the case of $\FPC$, the second-order variables may have mixed-type and we only
allow bounded quantification for number variables. We note additionally that
formulas may be treated as number terms that evaluate to zero or one.

We now briefly explain the intended semantics of the rank operator. Let $\phi
(\vec{x}, \vec {\nu}, \vec{y}\vec{\mu})$ be a number term, where $\vec{x}$ and
$\vec{y}$ are tuples of vertex variables, $\vec{\nu}$ and $\vec{\mu}$ are tuples
of number variables. Let and $\vec{s}$ and $\vec{t}$ be tuples of number terms
that bound the ranges of $\vec{\nu}$ and $\vec{\mu}$, respectively. Let $\pi$ be
a number variable and $r$ be a number term bounding the range of $\pi$. Let
$\mathcal{A}$ be a structure with universe $A$, and suppose we have a given
assignment to the all of the free variables appearing in $r$ and in the terms in
$\vec{t}$ and $\vec{s}$. Let $\mathbb{N}^{\leq\vec{t}} := \{\vec{n} \in
\mathbb{N}^{\vert \vec{v} \vert}: n_i \leq t^{\mathcal{A}}_i\}$, and let
$\mathbb{N}^{\leq \vec{s}}$ be defined similarly. Let $I:= A^{\vert \vec{x}
  \vert} \times \mathbb{N}^{\leq \vec{t}}$ and $J := A^{\vert \vec{y} \vert}
\times \mathbb{N}^{\leq \vec{s}}$. Let $M^{\mathcal{A}}_\phi : I \times J
\rightarrow \mathbb{N}$ be defined by $M^{\mathcal{A}}_\phi (\vec{a}\vec{n},
\vec{b}\vec{m}):= \phi^{\mathcal{A}}(\vec{a}\vec{n},\vec{b}\vec{m})$. For the
structure $\mathcal{A}$ and assignment $\pi \mapsto p$, for $p \leq r^{A}$
prime, $[\rank (\vec{x}, \vec {\nu} \leq \vec{t}, \vec{y}\vec{\mu} \leq \vec{s},
\pi \leq r). \phi]$ evaluates to the rank of $M^{\mathcal{A}}_\phi$ mod $p$,
where the entries in $M^{\mathcal{A}}$ are interpreted as elements of
$\mathbb{F}_p$.

We should note that there are two different of rank logics defined in the
literature. The first, introduced by Dawar et al.~\cite{Dawar09logicswith},
differs from the one defined above in that it does not take the characteristic
of the field as a parameter, instead including a different operator for every
prime field over which compute rank. It was shown by Grad{\"e}l and
Pakusa~\cite{GradelP15a} that this rank logic does not capture $\PT$. Grad{\"e}l
and Pakusa introduced the rank logic we have presented above and showed that
this logic, with the uniform rank operator, is strictly more expressive then the
rank logic introduced by Dawar et al. For this reason when we refer to rank
logic we are always referring to the rank logic introduced by Grad{\"e}l and
Pakusa~\cite{GradelP15a}.

We may also define an extension of first order logic by rank operators. Let
$\FOR[\tau]$ denote \emph{first-order with rank} with respect to the vocabulary
$\tau$, and be the set of formulas in $\FPR[\tau]$ that do not contain an
application of the fixed-point operator.

Let $\FPrk[\tau]$ denote \emph{fixed-point logic with rank quantifiers} with
respect to the vocabulary $\tau$. This logic, first introduced by Dawar et
al~\cite{Dawar09logicswith}, is similar to $\FPR$, except that it extends $\FP$
by \emph{rank quantifiers}, rather than a rank operator, that can decide if a
definable matrix has a rank bounded by some constant. We define for each prime
$p$, and natural number $r$, a rank quantifier $\rank^r_p$, such that
$\rank^r_p\vec{x}\vec{y}. \phi$ is interpreted as $[\rank (\vec{x}, \vec{y},
\pi) . \phi] \leq t_r$, where $\pi$ is assigned to $p$ and $t_r$ is a number
term that evaluates to $r$. The arity of this quantifier is $\vert \vec{x} \vert
+ \vert \vec{y} \vert$. Let $\mathcal{R}$ be the set of all such quantifiers and
$\FPrk[\tau]$ be the closure of $\FP[\tau]$ under $\mathcal{R}$.

We may also define an extension of first-order logic with rank quantifiers. Let
$\FOrk[\tau]$ denote \emph{first-order with rank quantifiers} with respect to
the vocabulary $\tau$, and be the set of all formulas in $\FPR[\tau]$ that do
not contain an application of the fixed-point operator.

For more detail on the syntax and semantics of logics rank quantifiers please
see~\cite{Dawar09logicswith}. It is easy to see that $\FPrk \leq \FPR$ and
$\FOrk \leq \FOR$. Moreover, we note that $[\rank(\vec{x}, \vec{y}).
(\vec{x} = \vec{y}, \pi \leq r) \land \phi(\vec{x})]$ evaluates, for a structure
$\mathcal{A}$ and any assignment $\pi \mapsto p$, to the number of assignments
to $\vec{x}$ for which $\phi$ holds (i.e.\ $\vert \{\vec{a} \in U^{\vert \vec{x}
  \vert} : \mathcal{A} \models \phi[\vec{a}]\} \vert$). It follows that $\FPC
\leq \FPR$ and $\FOC \leq \FOR$. Dawar et al.\cite{Dawar09logicswith} show that
all of these inclusions are strict.

@@

A \emph{Boolean basis} (or \emph{basis}) is a set of Boolean functions. We
always denote a basis by $\BB$. It is common in the literature to define a basis
using a finite sequence of functions $f_1, \ldots , f_k:\{0,1\}^* \rightarrow
\{0,1\}$. We write $\BB = \{f_1, \ldots, f_k\}$ to denote the basis consisting
of all Boolean functions $f^n_i : \{0,1\}^n \rightarrow \{0,1\}$ where $n \in
\nats$, $i \in [k]$, and $f^n_i$ is the restriction of $f_i$ to strings of
length $n$. We have $\AND, \OR, \NAND, \MAJ : \{0,1\}^* \rightarrow \{0,1\}$,
where $\AND, \OR, \NAND$ are the usual (unbounded fan-in) logical operators, amd
$\MAJ$ is the majority function. We write $\BS := \{\AND, \OR, \NOT\}$ to denote
the \emph{standard basis} and $\BS := \BS \cup \{\MAJ\}$ to denote the
\emph{majority basis}.

Let $\BB$ be a basis. A \emph{Boolean circuit} $C$ defined over $\BB$ is a
labelled directed acyclic graph (DAG) with designated input gates $x_1, \ldots,
x_n$, each with in-degree $0$, internal gates labelled by elements of $\BB$, and
a single internal gate with out-degree $0$ designated as the output gate. We say
a circuit with $n$ input gates has \emph{order $n$}. The evaluation of circuit
$C$ of order $n$ for the input $\vec{a} \in \{0,1\}^n$ is denoted by
$C[\vec{a}]$ and computed by recursively evaluation the gates in the circuit. We
denote the evaluation of a gate $g$ in $C$ for the input $\vec{a}$ by
$C[\vec{a}](g)$. If $g$ is an input gate, then $g = x_i$ for some $i \in [n]$,
and $C[\vec{a}](g) = a_i$. If $g$ is an internal gate then $g$ is labelled by a
symbol denoting a Boolean operation, and $C[\vec{a}](g)$ is the result of
applying that operation the string formed from the evaluations of those gates
input to $g$. If $g$ is the output gate, let $C[\vec{a}] = C[\vec{a}](g)$.

@@

\begin{thm}[Anderson and Dawar~\cite{AndersonD17}]
  For a relational vocabulary $\tau$, a property of finite $\tau$-structures is
  decided by a P-uniform family of symmetric circuits defined over the basis
  $\BS$ if, and only if, it is expressible in $\FPC[\tau]$.
\end{thm}

>>>>

Let $\FP[\tau]$ denote \emph{fixed-point logic} with respect to $\tau$. The
logic $\FP$ is an extension of $\FO$ by an inflationary fixed-point operator. In
other words, $\FP[\tau]$ includes all formulas in $\FO[\tau]$, and is closed
under applications of the fixed-pint operator, i.e. if $\phi \in \FP[\tau]$, $V$
is a second-order variable, and $\vec{x}$ is a tuple of variables of length
$\arty(V)$, then $[\ifp_{V, \vec{x}}. \phi] \in FP[\tau]$. We assume the
standard syntax and semantics for $\FO[\tau]$ and $\FP[\tau]$. For more detail
on the syntax and semantics of these logics please
see~\cite{grohe2017descriptive} and~\cite{GradelP15a}.

Let $L$ be a logic and $\tau$ a vocabulary. For $\phi \in L[\tau]$ and a
sequence of variables $\vec{x}$, we write $\phi(\vec{x})$ if the free variables
in $\phi$ are among those in $\vec{x}$. Suppose $\vec{x} = (x_1, \ldots , x_k)$,
and let $\mathcal{A}$ be a $\tau$-structure with universe $U$. We call a
function $\alpha : {x_1, \ldots, x_k} \rightarrow U$ an \emph{assignment}. If
$\vec{a}\in U^k$, we write $\alpha^{\vec{x}}_{\vec{a}}$ to denote the assignment
that maps the variable $x_i$ to $a_i$ for $i \in [k]$. We say that two
assignments $\alpha$ and $\beta$ are \emph{compatible} if $\alpha$ and $\beta$
agree on the intersection of their domain. We write $\mathcal{A} \models_L
\phi[\alpha]$ if, and only if, $\phi$ is true in $\mathcal{A}$ with respect to
$L$ where for all $i \in [k]$ each variable $x_i$ is assigned to $\alpha(x_i)$.
If the sequence of variables $\vec{x}$ that we are assigning to $\vec{a}$ is
obvious from context we write $\mathcal{A} \models_L \phi[\vec{a}]$ instead. If
the logic is obvious from context we omit the subscript and just write
$\mathcal{A} \models \phi[\alpha]$.

\subsection{Fixed-Point with Counting}
Let $\FPC[\tau]$ denote \emph{fixed-point logic with counting} with respect to
the vocabulary $\tau$. The logic $\FPC$ is an extension of $\FP$ that includes a
counting operator that allows us to define the size of a definable set, as well
as other mechanisms for reasoning about quantities. We now briefly describe the
structure of a formula of $\FPC[\tau]$. Every first-order variable is assigned a
type, and so each variable is either a \emph{vertex} or \emph{number} variable.
Vertex variables are assigned to elements of the universe of the structure, and
play a similar role to the variables that appear in formulas in $\FO[\tau]$ or
$\FP[\tau]$. Number variables are assigned to elements of $\nats$. We usually
use Latin letters (e.g. $x, y, z, \ldots$) to denote vertex variables and Greek
letters (e.g. $\mu, \nu, \epsilon, \ldots$) to denote number variables. All
atomic formulas in $\FP[\tau]$ are atomic formulas in $\FPC[\tau]$. We say that
$t$ is a \emph{number term} if $t$ is a number variable or if $t$ is an application
of the \emph{counting operator}, i.e. $t = \#x. \phi(x)$, where $\phi \in
\FPC[\tau]$ and $x$ is a vertex variable free in $\phi$. If $t_1$ and $t_2$ are
number terms then $t_1 \leq t_2$ and $t_1 = t_2$ are atomic formulas.
$\FPC[\tau]$ includes all formulas formed by closing the set of atomic formulas
under the usual Boolean connectives, the first-order quantifiers (for vertex
variables), and the inflationary fixed-point operator. The second-order
variables, such as those that appear in an application of the fixed-point
operator, are allowed to have mixed-type, i.e. we associate with a second-order
variable $R$ a type $(k, l) \in \nats^2$ intended to mean that $R$ is assigned
to a subset of $U^k \times \nats^l$.

@@

If $\mathcal{A}$ is a $\tau$-structure with universe $U$, the counting term
$\#x. \phi(x)$ evaluates to $\vert \{a \in U : \mathcal{A} \models \phi[a]\}
\vert$ in $\mathcal{A}$, assuming a fixed assignment to the other free variables
in $\phi$. The intended semantics for bounded quantification is as for ordinary
quantification but with the additional restriction that the assignment to the
number variable be bounded by the evaluation of the bounding number term. The
semantics of the `$=$' and `$\leq$' symbols are defined for number terms in the
obvious manner. For a more detailed description of the syntax and semantics of
$\FPC$ please see~\cite{grohe2017descriptive} or the introduction
of~\cite{GradelP15a}.

@@

Let $\FOC[\tau]$ denote \emph{first-order logic with counting} with respect to
the vocabulary $\tau$, and let $\FOC[\tau]$ be set of all formulas in
$\FPC[\tau]$ that do not contain an application of the fixed-point operator.

@@

\subsection{Rank Logic}
Let $\FPR[\tau]$ denote \emph{fixed-point logic with rank} (or just \emph{rank
  logic}) with respect to the vocabulary $\tau$. The logic $\FPR$ is an
extension of $\FP$ to include an operator that denotes the rank of a definable
matrix over a finite field, as well as other mechanisms for reasoning about
quantity. In this way we can think of the structure of a formula of $\FPR$ as
being similar to that of $\FPC$, with rank operators replacing counting
operators.

We now briefly describe the structure of a formula in $\FPR[\tau]$. For more
detail on the syntax and semantics of $\FPR$ please see~\cite{GradelP15a}
and~\cite{Dawar09logicswith}. As in the case of $\FPC$, each variable is either
a number or vertex variable, with vertex variables assigned to elements of the
universe and number variables assigned to natural numbers. All atomic formulas
in $\FP[\tau]$ are atomic formulas in $\FPR[\tau]$. We say that $t$ is a
\emph{number term} if $t$ is a number variable or if $t$ is an application of
the \emph{rank operator}, i.e. $t = [\rank (\vec{x}, \vec {\nu} \leq \vec{t},
\vec{y}\vec{\mu} \leq \vec{s}, \pi \leq \eta). \phi]$, where $\phi$ is a number
term, $\vec{t}$ and $\vec{s}$ are tuples of number terms bounding the sequences
of number variables $\vec{\mu}$ and $\vec{\nu}$, and $\eta$ is a number term
bounding the number variable $\pi$. If $t_1$ and $t_2$ are number terms then
$t_1 \leq t_2$ and $t_1 = t_2$ are atomic formulas. The formulas of $\FPR$ are
then formed by closing the set of atomic formulas under the usual Boolean
connectives, the first-order quantifiers, and the fixed-point operator. As in
the case of $\FPC$, the second-order variables may have mixed-type and we only
allow bounded quantification for number variables. We note additionally that
formulas may be treated as number terms that evaluate to zero or one.

We now briefly explain the intended semantics of the rank operator. Let $\phi
(\vec{x} \vec {\nu}, \vec{y}\vec{\mu})$ be a number term, where $\vec{x}$ and
$\vec{y}$ are tuples of vertex variables, $\vec{\nu}$ and $\vec{\mu}$ are tuples
of number variables. Let $\vec{s}$ and $\vec{t}$ be tuples of number terms such
that $\vert \vec{s} \vert = \vert \vec{\nu} \vert$ $\vert \vec{t}\vert = \vert
\vec{mu} \vert$. Let $\pi$ be a number variable and $r$ be a number term. Let
$\mathcal{A}$ be a structure with universe $A$, and suppose we have a given
assignment to the those free variables appearing in $r$ or in the terms in
$\vec{t}$ and $\vec{s}$. Let $\mathbb{N}^{\leq\vec{t}} := \{\vec{n} \in
\mathbb{N}^{\vert \vec{\nu} \vert}: n_i \leq t^{\mathcal{A}}_i\}$, and let
$\mathbb{N}^{\leq \vec{s}}$ be defined similarly. Let $I:= A^{\vert \vec{x}
  \vert} \times \mathbb{N}^{\leq \vec{t}}$ and $J := A^{\vert \vec{y} \vert}
\times \mathbb{N}^{\leq \vec{s}}$. Let $M^{\mathcal{A}}_\phi : I \times J
\rightarrow \mathbb{N}$ be defined by $M^{\mathcal{A}}_\phi (\vec{a}\vec{n},
\vec{b}\vec{m}):= \phi^{\mathcal{A}}(\vec{a}\vec{n},\vec{b}\vec{m})$. For the
structure $\mathcal{A}$ and assignment $\pi \mapsto p$, where $p \leq
r^{\mathcal{A}}$, if $p$ is prime then the rank term $[\rank (\vec{x}, \vec
{\nu} \leq \vec{t}, \vec{y}\vec{\mu} \leq \vec{s}, \pi \leq r). \phi]$ is the
rank of $M^{\mathcal{A}}_\phi$ mod $p$, where the entries in $M^{\mathcal{A}}$
are interpreted as elements of $\mathbb{F}_p$. If $p$ is not prime then the rank
term evaluates to zero.

We should note that there are two distinct rank logics defined in the
literature. The first was introduced by Dawar et al.~\cite{Dawar09logicswith},
and differs from the rank logic defined above in that it includes a sequence of
rank operator, rather than a single uniform operator, one for each prime $p$
expressing the rank over the field $\ff_p$. It was shown by Grad{\"e}l and
Pakusa~\cite{GradelP15a} that this non-uniform rank logic does not capture
$\PT$. After proving this result, Grad{\"e}l and Pakusa introduced the rank
logic with a uniform rank operator that we have presented above, and showed that
this rank logic is strictly more expressive then the rank logic introduced by
Dawar et al. This uniform rank logic remains a candidate for capturing
polynomial-time. For this reason when we refer to rank logic in this paper we
are always referring to the rank logic introduced by Grad{\"e}l and
Pakusa~\cite{GradelP15a}.

We may also define an extension of first order logic by rank operators. Let
$\FOR[\tau]$ denote \emph{first-order with rank} with respect to the vocabulary
$\tau$, and let $\FOR[\tau]$ be the set of formulas in $\FPR[\tau]$ that do not
contain an application of the fixed-point operator.

Let $\FPrk[\tau]$ denote \emph{fixed-point logic with rank quantifiers} with
respect to the vocabulary $\tau$. This logic, first introduced by Dawar et
al~\cite{Dawar09logicswith}, is similar to $\FPR$, except that it extends $\FP$
by a set of \emph{rank quantifiers}, rather than a rank operator. We define for each prime
$p$, and natural number $r$, a rank quantifier $\rank^r_p$, such that
$\rank^r_p\vec{x}\vec{y}. \phi$ is interpreted as $[\rank (\vec{x}, \vec{y},
\pi) . \phi] \leq t_r$, where $\pi$ is assigned to $p$ and $t_r$ is a number
term that evaluates to $r$. The arity of this quantifier is $\vert \vec{x} \vert
+ \vert \vec{y} \vert$. Let $\mathcal{R}$ be the set of all such quantifiers and
$\FPrk[\tau]$ be the closure of $\FP[\tau]$ under $\mathcal{R}$.

We may also define an extension of first-order logic with rank quantifiers. Let
$\FOrk[\tau]$ denote \emph{first-order with rank quantifiers} with respect to
the vocabulary $\tau$, and let $\FOrk[\tau]$ be the set of all formulas in
$\FPR[\tau]$ that do not contain an application of the fixed-point operator.

For more detail on the syntax and semantics of logics with rank quantifiers please
see~\cite{Dawar09logicswith}. It is easy to see that $\FPrk \leq \FPR$ and
$\FOrk \leq \FOR$. Moreover, we note that $[\rank(\vec{x}, \vec{y}). (\vec{x} =
\vec{y}, \pi \leq r) \land \phi(\vec{x})]$ evaluates, for a structure
$\mathcal{A}$ and any assignment $\pi \mapsto p$, to the number of assignments
to $\vec{x}$ for which $\phi$ holds (i.e.\ $\vert \{\vec{a} \in U^{\vert \vec{x}
  \vert} : \mathcal{A} \models \phi[\vec{a}]\} \vert$). It follows that $\FPC
\leq \FPR$ and $\FOC \leq \FOR$. Dawar et al.\cite{Dawar09logicswith} show that
all of these inclusions are strict.

@@

A \emph{Boolean basis} (or \emph{basis}) is a set of Boolean functions. We
always denote a basis by $\BB$. It is common in the literature to define a basis
using a finite sequence of functions $f_1, \ldots , f_k:\{0,1\}^* \rightarrow
\{0,1\}$. We write $\BB = \{f_1, \ldots, f_k\}$ to denote the basis consisting
of all Boolean functions $f^n_i : \{0,1\}^n \rightarrow \{0,1\}$ where $n \in
\nats$, $i \in [k]$, and $f^n_i$ is the restriction of $f_i$ to strings of
length $n$. We have $\AND, \OR, \NAND, \MAJ : \{0,1\}^* \rightarrow \{0,1\}$,
where $\AND, \OR, \NAND$ are the usual (unbounded fan-in) logical operators, and
$\MAJ$ is the majority function. We write $\BS := \{\AND, \OR, \NOT\}$ to denote
the \emph{standard basis} and $\BM := \BS \cup \{\MAJ\}$ to denote the
\emph{majority basis}.

Let $\BB$ be a basis. A \emph{Boolean circuit} $C$ defined over $\BB$ is a
labelled directed acyclic graph (DAG) with designated input gates $x_1, \ldots,
x_n$, each with in-degree $0$, internal gates labelled by elements of $\BB$, and
a single internal gate with out-degree $0$ designated as the output gate. We say
a circuit with $n$ input gates has \emph{order $n$}. The evaluation of circuit
$C$ of order $n$ for the input $\vec{a} \in \{0,1\}^n$ is denoted by
$C[\vec{a}]$ and computed by recursively evaluating the gates in the circuit. We
denote the evaluation of a gate $g$ in $C$ for the input $\vec{a}$ by
$C[\vec{a}](g)$. If $g$ is an input gate, then $g = x_i$ for some $i \in [n]$,
and $C[\vec{a}](g) = a_i$. If $g$ is an internal gate then $g$ is labelled by a
symbol denoting a Boolean operation, and $C[\vec{a}](g)$ is the result of
applying that operation to the string formed from the evaluations of those gates
input to $g$. If $g$ is the output gate, let $C[\vec{a}] = C[\vec{a}](g)$.

@@

\begin{thm}[Anderson and Dawar~\cite{AndersonD17}]
  For a relational vocabulary $\tau$, a property of finite $\tau$-structures is
  decided by a P-uniform family of symmetric circuits defined over the basis
  $\BM$ if, and only if, it is expressible in $\FPC[\tau]$.
\end{thm}

\subsection{Equivalence Relations}
Let $\sim$ be an equivalence relation on the set $X$. For $x \in X$ we write
$[x]_\sim$ to denote the equivalence class of $x$ with respect to $\sim$. We
omit the subscript if the equivalence relation is obvious from context. We write
$X/_{\sim}$ to denote the set of equivalence classes of $X$. Let $A = \Pi_{i \in
  I} A_i$ and let $J \subseteq I$ be such that for $i \in J$, $A_i \subseteq X$.
Let $\sim_A$ be an equivalence relation on $A$ defined such that $\vec{a} \sim_A
\vec{b}$ if, and only if, $\vec{a}(i) = \vec{b}(i)$ for all $i \in J$. We call
$\sim_A$ the \emph{extension} of $\sim$ to $A$. We often abuse notation and
write $\sim$ for the extension of the equivalence relation $\sim$. We notice
that $(\Pi_{i \in I} A_i) /_{\sim} = \Pi_{i \in I} (A_i /_\sim)$.

Let $Y$ and $Z$ be sets and let $f : Y \rightarrow Z$ be a function. We say $f$
is \emph{invariant} under $\sim$ if for all $a, b \in Y$ if $a \sim_Y b$ then
$f(a) \sim_Z f(b)$. If $f$ is an invariant under $\sim$ then $f /_{\sim}$ is a
function mapping $Y /_{\sim_Y} $ to $ Z /_{\sim_Z}$ such that $f /_{\sim} ([a]) =
[f(a)]$ for all $a \in Y$. We call $f_{\sim}$ the \emph{quotient} of $f$ with
respect to $\sim$.
